---
title: "Tree/Heatmap Plots"
author: "John Quensen"
date: "4/21/2021"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

This script generates a tree/heatmap plot for Jose's 286 genomes from clinical and cattle samples.  

This script details how to make a tree plot based on  ANI distances among genomes and align it to a heatmap of orthologous genes. This aids in interpretation of the tree structure.   

Load packages.  

```{r, tidy=TRUE}
suppressPackageStartupMessages(library(ape))
suppressPackageStartupMessages(library(phangorn))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(hablar))
suppressPackageStartupMessages(library(ggtree))
suppressPackageStartupMessages(library(aplot))
```

Read in data. Edit the path to the MiGA project in the first line of the next code block. The other paths are those within MiGA's output directory structure.  

```{r, tidy=TRUE}
# Read in the ANI tree file.
ani.tree <- ape::read.tree("extra_data/jose-miga-project.ani.nwk")

# Read in the orthologous genes file.
ogs <- read.table("extra_data/jose-miga-project.ogs",
                  header = TRUE,
                  stringsAsFactors = FALSE,
                  sep = "\t")
```

`ogs` is a text file giving the genomes in columns and orthologs in rows. A sample is:  

```{r, tidy=TRUE}
ogs[1:5, 1:5]
```

The cell entries are the gene IDs as assigned by MiGA.    

Convert `ogs` to a binary (presence/absence) matrix. Assign arbitrary row names to serve as the ortholog names.   

```{r, tidy=TRUE}
ogs.bin <- ogs
ogs.bin[ogs.bin!="-"] <- 1
ogs.bin[ogs.bin=="-"] <- 0
ogs.bin[1:5, 1:5]
rownames(ogs.bin) <- RDPutils::make_otu_names(1:nrow(ogs.bin))
rownames(ogs.bin) <- sub("OTU", "OG", rownames(ogs.bin))
ogs.bin[1:5, 1:4]
```

Convert to a tibble for making the heatmap plot.  

```{r, tidy=TRUE}
n.cols <- ncol(ogs.bin) + 1
df.ogs <- ogs.bin %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "Gene") %>% 
  hablar::convert(int(colnames(.)[2:n.cols])) %>% 
  hablar::convert(fct(Gene)) %>% 
  tidyr::pivot_longer(-Gene, names_to = "Genome") %>% 
  tibble::as_tibble()

df.ogs
```

Make the heatmap plot.  

```{r, tidy=TRUE}
heatmap.plt <- ggplot(df.ogs, aes(x = Gene, y = Genome, fill = as.factor(value))) +
  geom_tile() +
  scale_fill_manual(values = c("white", "blue")) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.line.x.bottom = element_line(size = 0.5),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(legend.position = "none") +
  ylab(NULL)

# heatmap.plt
```

Root the tree. Tree tip labels and rows in the heatmap cannot be aligned with `aplot` unless the tree is rooted. A simple solution is to use `phanghorn`'s `midpoint` function.    

```{r, warning=FALSE, message=FALSE, tidy=TRUE}
ani.tree <- phangorn::midpoint(ani.tree)
ani.tree
```

Make a plot of the tree. Expand the plot so that the tip labels show.   
 
```{r, tidy=TRUE}
gg_tr <- ggtree(ani.tree) +
  geom_tiplab(align=TRUE, size=0.5) +
  scale_x_continuous(expand=expansion(0,0.001))
# gg_tr
```

Align the plot by the tip labels, side-by-side. The `aplot` package automatically sizes the plots so that the tree tips and heatmap rows align. The plot can be saved with the `ggsave` function.   

```{r, tidy=TRUE}
plt <-  heatmap.plt %>% insert_left(gg_tr)
# plt
ggsave(plt, file = "jose-tree_heatmap.png", width = 7, height = 7, units = "in", dpi = 600)
```

Make the same plot, but color the tip labels according the the data source: cattle or clinical samples.   

```{r}
tip.data <- as.data.frame(ani.tree$tip.label)
colnames(tip.data) <- 'label'
tip.data <- tip.data %>% 
  mutate(study = if_else(grepl("T", label), "Clinical", "Cattle"))
head(tip.data)

gg_tr <- ggtree(ani.tree)

gg_tr <- gg_tr %<+% tip.data +
  geom_tiplab(aes(color = study), align=TRUE, size=0.5) +
  scale_x_continuous(expand=expansion(0,0.001)) +
  theme(legend.position = 'none')

plt <-  heatmap.plt %>% insert_left(gg_tr)
# plt
ggsave(plt, file = "jose-tree_heatmap_tips.png", width = 7, height = 7,
       units = "in", dpi = 600)
```

